Associations of environmental and human factors with altered rodent populations in West Africa: A scoping review

### 2021-04-10 reports/manuscript.Rmd
caught up with new data

A total of 2,200 records were identified, with 101 studies included in narrative synthesis (see Figure \@ref(fig:prisma)). The characteristics of the included studies are shown in Table 1. The earliest studies identified were from 1974 with the majority (62%) published since 2010. Most of the included studies were classified as rodent ecology research (54%) with the remainder of studies on zoonotic diseases (46%).

@@

The detailed aims of the included studies are shown following division into the themes of ecological or zoonosis research. Biodiversity of rodents was a shared aim between the two themes with 37 ecological studies and 2 zoonoses studies investigating this. The remaining aims were more disparate with ecological studies investigating rodent population dynamics, rodent species assemblages, the genetics of rodents and their behaviours. Most zoonotic disease research focussed on prevalence of pathogens in rodent populations, the risk of spillover to human populations, the identity of pathogen reservoir species with fewer studies investigating the phylogenetics of pathogen strains (see Figure \@ref(fig:aims).

@@

The included studies trapped in 16 countries, however, three countries were outside of the scope of this study (Cameroon, Chad and Morocco). At least a single trap site was recorded from 13 West African countries, with no rodent trapping reported from Gambia, Cape Verde or Togo.

Most studies (95%) were conducted in a single country, three studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries. This lead to 124 trapping activities being reported across the 13 West African countries. More trapping activity was reported from Senegal (20%) than any other country. Ghana (12%) and Guinea (12%) were the next most frequently trapped countries, followed by Nigeria (9%), Benin, Mali and Niger (7%), Sierra Leone (6%), Mauritania and the Ivory Coast (4%), Burkina Faso and Liberia (2%), Guinea-Bissau had a single recorded trapping activity.

The trapping activity from the included studies was spread across 1,171 study sites. Twenty-four (24%) studies reported trapping only occurring at a single study site, 34 (33%) studies trapped at between two and five study sites the remaining 44 studies trapped at between six and 93 study sites. Trap sites were mapped to level one country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)). The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast. Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore) and Eastern Sierra Leone were most intensively studied. Mapping to level two administrative regions is displayed in Supplementary figure \@ref(fig:supfig1).

@@

Of the included studies 15 did not state the type of trap used with three studies reporting supplementary hand trapping of rodents alongside trap use. For the remaining studies three used more than four types of rodent traps, seven used three types, 31 used two types the remaining 45 studies used a single trap type. The most commonly used trap was the Sherman trap (67%) [ref.] which is available in several sizes, locally made wire mesh traps were used in 29 studies (34%), less commonly used trap types included the Museum special, Tomahawk, Victor, Snaptraps and Firobind traps.

@@

Thirty-three studies did not describe the structure of the trap setup within the defined study area. Twenty five studies placed transect lines of traps through study sites with thirteen studies placing traps along a line but not clearly describing a transect across a defined site. Sixteen studies purposefully sampled houses, with four further studies purposefully placing traps in targeted locations (i.e. near rivers, in fields, by houses). Nine studies placed traps in a structured grid across a single or multiple specified habitat types. Three studies described a stratified or randomised approach to placing traps within a defined area. Trap setup was commonly reported in ecology studies (80%) with transect lines the most commonly used approach. Trap setup was reported in 53% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (20%).

@@

Trapping effort was variably reported across included studies. Thirty studies did not report any measures of trapping effort. Thirty-four studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classified as complete reports. Thirty-seven studies had incomplete reports on trapping effort due to not presenting the number of trap nights at the appropriate unit of analysis (i.e. number of study nights only or total number of trap-nights). Trapping effort was more completely reported from ecology studies (51% complete information and 20% no information) compared to zoonoses risk studies (13% complete information and 41% no information).

Among studies that reported complete trap-night information the median number of trap-nights per study was 4,764 (Interquartile range (IQR) = 1,227-10,477) with a minimum of 240 trap-nights and maximum of 45,274 trap-nights. For studies reporting the number of trap-nights at the unit of analysis (typically study site or habitat type) the median number of trap-nights was 360 (IQR = 246-793). In studies with incomplete reports on trap-nights at the unit of analysis there was a median of 7,893 trap-nights (IQR = 1,401-10,331) conducted, for 21 studies reporting only on the number of days or nights trapping was performed on activities typically lasted between 1-4 nights.

@@

No studies classified habitats with reference to a specific habitat classification scheme. Seven studies did not describe the local habitat where traps were placed (3 ecology, 4 zoonoses). The remaining studies used greater than 160 descriptions of the habitat type in which trapping was conducted, further, the resolution of the description ranged from the habitat of the location of the trap through to the study site. The 160 reported habitat types were combined into 32 categories (see Supplementary table 2 for the habitat dictionary) for further analysis. At least one habitat was recorded for 16,168 trap sites (95%), with two or more habitats for a single trap site recorded for 4,195 (26%) sites. Single trap sites could span multiple habitat types resulting in 20,369 habitat or trap sites. The most commonly trapped sites were in or around buildings (31%), in areas described as the rodents "natural habitat" (26%), in agricultural areas (13%) (e.g. rice fields, palm plantations), forests (6%) and within or around villages (5%).

Habitat type was more comprehensively reported from ecology studies, 31 habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), in or around buildings (18%) and within forests (15%). Among studies on zoonoses 15 habitat categories were reported, the most commonly reported classification was described as the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).

@@

Nine included studies did not describe the method used to identify trapped rodents. Most studies (52%) relied solely on morphometric identification against taxonomic keys or museum voucher specimens. Eight studies described molecular identification of rodents with 30% using a combination of morphological and molecular techniques. In more recently published studies (since 2010) 54% used molecular techniques to assist with speciation, compared with less than 16% in studies published before 2010.

@@

Across all West African study sites there were 59,343 trapped small mammals (592 were trapped outside of West African countries), 2,830 (4.8%) trapped individuals were identified to order level (Rodentia), 5,329 (9%) were identified to genus level, the remaining 51,184 (86%) were identified to species level. Of the 124 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (96) with Muridae (79) being the largest family of rodents, followed by Gliridae and Nesomyidae (both 4), Sciuridae (3), Ctenodactylidae (2) and Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (all 1). The remaining 29 species came from the orders of the Soricomorpha (25), Erinaceomorpha (2) and Afrosoricida (1).

More individuals were trapped in studies conducted to investigate the risk of zoonoses (35,892, 60%) than from ecology studies (23,451 40%). Animals trapped as part of ecological research were more often identified to species level (93%) than those trapped in zoonoses research (82%). The most commonly trapped genera of rodents were *Mastomys sp.* (22,665, 39.7%), *Rattus sp.* (7,718, 13.5%), *Mus sp.* (7,324, 12.8%), *Praomys sp.* (4,977, 8.7%) and *Arvicanthis sp.* (4,408, 7.72%). At the species level *Mastomys natalensis* (10,351, 20.2%) and *Mastomys erythroleucus* (7,052, 13.8%) were the most commonly trapped rodents, *Rattus rattus* (6,228, 12.2%), *Mus musculus* (6,030, 11.8%), *Arvicanthis niloticus* (4,086, 8%) and *Mastomys huberti* (3,106, 6.1%) made up the next most commonly trapped species. Presence and absence maps for these six most commonly trapped species are shown in Figure \@ref(fig:dist).

>>>>

A total of 4,282 records were identified, with 124 studies included in narrative synthesis (see Figure \@ref(fig:prisma)). The characteristics of the included studies are shown in Table 1. The earliest studies identified were from 1974 with the majority (61%) published since 2010. Most of the included studies were classified as rodent ecology research (56%) with the remainder of studies on zoonotic diseases (44%).

@@

The detailed aims of the included studies are shown in Figure \@ref(fig:aims). Biodiversity of rodents was a shared aim between the two themes with 37 ecological studies and 2 zoonoses studies investigating this. The remaining aims were more disparate with ecological studies investigating rodent population dynamics, rodent species assemblages, the genetics of rodents and their behaviours. Most zoonotic disease research focussed on prevalence of pathogens in rodent populations, the risk of spillover to human populations, the identity of pathogen reservoir species with fewer studies investigating the phylogenetics of pathogen strains.

@@

The included studies trapped in 17 countries, however, three countries were outside of the scope of this study (Cameroon, Chad and Morocco). At least a single trap site was recorded from 14 West African countries, with no rodent trapping reported from Gambia or Togo.

Most studies (93%) were conducted in a single country, five studies were conducted in two countries, three studies were conducted in five countries and one study was conducted across six countries. This lead to 146 trapping activities being reported across the 14 West African countries. More trapping activity was reported from Senegal (19%) than any other country. Nigeria (13%), Ghana and Guinea (both 12%) were the next most frequently trapped countries, followed by Mali and Benin (both 8%), Niger and Sierra Leone (both 6%), Mauritania and the Ivory Coast (4%), Burkina Faso (3%), Liberia (2%) with Guinea-Bissau and Cape Verde recording a single trapping activity.

The trapping activity from the included studies was spread across 1,193 study sites. Thirty-one (25%) studies reported trapping only occurring at a single study site, 45 (37%) studies trapped at between two and five study sites the remaining 48 studies trapped at between six and 93 study sites. Trap sites were mapped to level two country administrative regions and density of sites per million square kilometers was calculated (see Figure \@ref(fig:density)). The areas with highest trapping densities included the capital cities of Niger, Sierra Leone, Senegal, Benin, Mali, Ghana and the Ivory Coast. Of the areas not incorporating capital cities Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were most intensively studied.

@@

Of the included studies 18 did not state the type of trap used with three studies reporting supplementary hand trapping of rodents alongside trap use. For the remaining studies three used more than four types of rodent traps, twelve used three types, 33 used two types the remaining 58 studies used a single trap type. The most commonly used trap was the Sherman trap (56%) which is available in several sizes, locally made wire mesh traps were used in 35 studies (28%), less commonly used trap types included the Museum special, Tomahawk, Victor, Snaptraps and Firobind traps.

@@

Forty studies did not describe the structure of the trap setup within the defined study area. Twenty seven studies placed transect lines of traps through study sites with nineteen studies placing traps along a line but not clearly describing a transect across a defined site. Twenty-two studies purposefully sampled houses, with nine further studies purposefully placing traps in targeted locations (i.e. near rivers, in fields, by houses). Ten studies placed traps in a structured grid across a single or multiple specified habitat types. Three studies described a stratified or randomised approach to placing traps within a defined area. Trap setup was commonly reported in ecology studies (81%) with transect lines the most commonly used approach. Trap setup was reported in 56% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (24%).

@@

Trapping effort was variably reported across included studies. Thirty-nine studies did not report any measures of trapping effort. Forty-two studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classified as complete reports. Thirty-nine studies had incomplete reports on trapping effort due to not presenting the number of trap nights at the appropriate unit of analysis (i.e. number of study nights only or total number of trap-nights). Trapping effort was more completely reported from ecology studies (51% complete information and 21% no information) compared to zoonoses risk studies (11% complete information and 44% no information).

Among studies that reported complete trap-night information the median number of trap-nights per study was 4,418 (Interquartile range (IQR) = 1,187-10,252) with a minimum of 240 trap-nights and maximum of 45,274 trap-nights. For studies reporting the number of trap-nights at the unit of analysis (typically study site or habitat type) the median number of trap-nights at each site was 360 (IQR = 246-793). In studies with incomplete reports on trap-nights there was a median of 7,200 trap-nights (IQR = 1,184-9,424) conducted, for 26 studies reporting only on the number of days or nights trapping was performed on activities typically lasted between 1-4 nights.

@@

No studies reported trap habitats with direct reference to a habitat classification scheme. Thirteen studies did not describe the local habitat where traps were placed (4 ecology, 9 zoonoses). The remaining studies used more than 170 descriptions of the habitat type in which trapping was conducted, further, the resolution of the description varied between studies with some reporting the habitat of the location of the trap and others the study site or region. The reported habitat types were combined into 30 categories (see Supplementary table 2 for the habitat dictionary) for further analysis. At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites. Single trap sites could span multiple habitat types resulting in 22,202 habitat or trap sites. The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents "natural habitat" (24%), in agricultural areas (13%) (e.g. rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).

Habitat type was more comprehensively reported from ecology studies, 30 habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), within forests (18%) and in or around buildings (16%). Among studies on zoonoses 15 habitat categories were reported, the most commonly reported classification was described as the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).

@@

Thirteen included studies did not describe the method used to identify trapped rodents. Most studies (52%) relied solely on morphometric identification against taxonomic keys or museum voucher specimens. Nine studies described molecular identification of rodents with 30% using a combination of morphological and molecular techniques. In more recently published studies (since 2010) 76% used molecular techniques to assist with speciation, compared with less than 15% in studies published before 2010.

@@

Across all West African study sites there were 73,164 trapped small mammals (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level. Of the 147 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (112) with Muridae (82) being the largest family of rodents, Sciuridae (10), followed by Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1). The remaining 34 species came from the orders of the Soricomorpha (31), Erinaceomorpha (2) and Afrosoricida (1).

A similar number of rodents were trapped for ecology studies (36,367) as for studies investigating the risk of zoonoses (36,205). Animals trapped as part of ecological research were more often identified to species level (90%) than those trapped in zoonoses research (82%). The most commonly trapped genera of rodents were *Mastomys sp.* (27,072, 38.5%), *Rattus sp.* (10,191, 14.5%), *Mus sp.* (8,624, 12.3%), *Arvicanthis sp.* (5,821, 8.3%) and *Praomys sp.* (5,409, 7.7%). At the species level *Mastomys natalensis* (11,221, 17.9%) and *Rattus rattus* (8,578, 13.7%) were the most commonly trapped rodents, *Mastomys erythroleucus* (8,578, 13.7%), *Mus musculus* (6,245, 8.8%), *Arvicanthis niloticus* (5,497, 8.8%) and *Mastomys huberti* (4,699, 7.5%) made up the next most commonly trapped species. Presence and absence maps for these six most commonly trapped species are shown in Figure \@ref(fig:dist).
